---
title: "To Test, or Not to Test?"
subtitle: "A Hypothetical Report for Georgecity University"
authors:
  - name: Kylie King
    email: kk1543@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
bibliography: reference.bib
title-block-banner: '#27b9f8ff'
title-block-banner-color: "black"
format:
  html:
    df-print: kable
  pdf:
    # https://quarto.org/docs/reference/formats/pdf.html
    link-citations: false
prefer-html: true
---

# Introduction
The “Scholastic Assessment Test,” or the SAT, is a major part of the college application process. It is a college entrance exam that is meant to provide a uniform metric of a student's preparedness for college. Before 2021, most colleges and universities required prospective students to submit an SAT score (or it's counterpart, the American College Test) as part of their application package. However, in the aftermath of the COVID-19 pandemic, many colleges implemented test-optional policies, waiving the testing requirement for prospective students. In 2026, six years after the onset of the pandemic, many students, parents, and institutions are wondering: Is it time to reinstate the SAT?
(cite stuff)

## The Scholastic Assessment Test
The SAT is a standardized test that is typically taken in high school, in preparation for the college application process. It scores students on a scale of 200-1600 in Math, Reading/Writing, and optional Essay sections. College Board, the organization that administers the SAT, claims that students who score at least a 480 in Reading/Writing and 530 in Math, or a 1010 overall, are on track to be 'college ready'.[@sat_benchmarks]

College Board provides the following statement about the purpose and capabilities of the SAT:

> The primary purpose of the digital SAT Suite is to assess students’ preparedness for success in both college and careers. All assessment content is developed based on high-quality research that identifies the knowledge and skills essential for college and career readiness and success. Each test within the suite is designed to collect evidence from student performance to support broad claims about what students know and can do. This aligns with the primary purpose of assessing readiness for college and career training programs. [@sat_purpose]

To put it more simply, the SAT is designed to assess how well a student will perform in college. College Board published a study in 2023, which claims that “SAT scores add, on average, 15% more predictive power than high school GPA alone in understanding student performance in the first year of college." [@sat_study] It is easier for a college to determine how well a student will perform at their institution when they have the SAT as a metric. Being such a large part of the college application process, the SAT is a source of stress for many students, and some believe it creates a class barrier for low-income students who cannot afford the test, or its preparation. [@Kantrowitz_2022]


## The Introduction of Test-Optional Policies
During the COVID-19 pandemic, it became extremely difficult for students to take the SAT. Many test centers were shut down due to local states of emergency. In addition, many students faced health concerns, financial constraints, or mental health crises that prevented them from taking the exams. In a response to the pandemics restrictions, many schools implemented 'test-optional' policies. Under these policies, the schools did not require students to take the SAT (or ACT) to apply for undergraduate admission. As a result, the number of students who took the SAT in 2021 was 30% lower than in 2020, and that number has not returned to pre-pandemic levels as of 2024's published metrics. [@USAFacts_Team_2022] This decline of test-takers over time is displayed in the line chart below, to show the effect of the COVID-19 pandemic on SAT testing.

```{r}
#| label: imports
#| echo: false
#| warning: false
#| message: false
library(tidyverse, quietly = TRUE, warn.conflicts = FALSE) |> suppressPackageStartupMessages()
library(plotly, quietly = TRUE) |> suppressPackageStartupMessages()
library(colorspace, quietly = TRUE) |> suppressPackageStartupMessages()
dark_steelblue <- darken("steelblue4", 0.2)
sat_blue <- '#009bdb'
sat_grey <- "#96999aff"

```

```{r}
#| label: loading_data
#| echo: false
admin_df <- read_csv("data/cleaned_admin_df.csv", show_col_types = FALSE)
admin_test_df <- read_csv("data/admin_test_full.csv", show_col_types = FALSE)
df_counts <- admin_test_df |>
  count(Policy_2026) |>
  mutate(pct = n / sum(n) * 100)
```


```{r}
#| echo: false
year <- c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
test_takers <- c(1637589, 1715481, 2136539, 2220087, 2198460, 1509133, 1737678, 1913742, 1973891)
test_takers_df <- data.frame(year, test_takers)

fig <- plot_ly(
  test_takers_df,
  x = ~year,
  y = ~test_takers/1e6,
  type = 'scatter',
  mode = 'lines+markers',
  line = list(color = sat_blue, width = 2),
  text = ~paste(
    "Year: ", year,
    "<br>Number of Students:", format(test_takers, big.mark = ",")
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "SAT Test Takers by Year (2016-2024)",
      x = 0.5,
      font = list(size = 16, color = "black", family = "Arial") 
    ),
    xaxis = list(
      title = list(text="Year", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text="Number of Test Takers (millions)", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".1f",
      fixedrange = TRUE
    )
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
#| echo: false
fig
```





```{r}
#| label: testing charts
#| echo: false
#testing plotly charts

fig <- plot_ly(admin_df, 
               x = ~SAT_AVG, y = ~C150_4,
               type = 'scatter', mode = 'markers',
               marker = list(color = sat_blue),
               name = 'Schools',
               text = ~paste(
    "School: ", INSTNM,
    "<br>Average SAT: ", SAT_AVG,
    "<br>Completion Rate: ", scales::percent(C150_4, accuracy = 1)
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "Relationship Between SAT Average Scores and 6-Year Completion Rates",
      font = list(size = 18, color = "black", family = "Arial"),
      x = 0.5  # centers the title like hjust = 0.5
    ),
    margin = list(t = 100),
    xaxis = list(
      title = list(text = "SAT Average Score for Accepted Students", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text = "Undergraduate 6-Year Completion Rate", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".0%",
      fixedrange = TRUE
    ),
    legend = list(
      title = list(text = "", font = list(size = 14, color = "black")),
      font = list(size = 12, color = "black")
    ),
    dragmode = "zoom" 
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
#| echo: false
fig
```

```{r}
#| echo: false
fig <- plot_ly(
  df_counts,
  x = ~Policy_2026,
  y = ~pct/100,
  type = 'bar',
  marker = list(
    color = ifelse(df_counts$Policy_2026 == "Test Optional", sat_blue, sat_grey)
  ),
  text = ~paste0(round(pct, 1), "%"),   # Text labels as percentages
  textposition = 'outside',              # Place text above bars
  textfont = list(size = 14, color = "black", family = "Arial")  # Match scatterplot style
) |>
  layout(
    title = list(
      text = "Admissions Test Policies in Nearby Schools for Fall 2026",
      x = 0.5,
      font = list(size = 18, color = "black", family = "Arial")  # Match scatterplot title style
    ),
    xaxis = list(
      title = list(text="Admissions Testing Requirement", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial")
    ),
    yaxis = list(
      title = list(text="Percent of Nearby Schools", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".0%",
      range = c(0, 1)
    ),
    margin = list(t = 90)  # Extra top margin for title
  ) |>
  config(
    staticPlot = TRUE 
  )

```

```{r}
#| echo: false
fig
```

```{r}
#| echo: false

year <- c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
test_takers <- c(1637589, 1715481, 2136539, 2220087, 2198460, 1509133, 1737678, 1913742, 1973891)
test_takers_df <- data.frame(year, test_takers)

fig <- plot_ly(
  test_takers_df,
  x = ~year,
  y = ~test_takers/1e6,
  type = 'scatter',
  mode = 'lines+markers',
  line = list(color = sat_blue, width = 2),
  text = ~paste(
    "Year: ", year,
    "<br>Number of Students:", format(test_takers, big.mark = ",")
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "SAT Test Takers by Year (2016-2024)",
      x = 0.5,
      font = list(size = 16, color = "black", family = "Arial") 
    ),
    xaxis = list(
      title = list(text="Year", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text="Number of Test Takers (millions)", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".1f",
      fixedrange = TRUE
    )
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
#| echo: false
fig
```