---
title: "To Test, or Not to Test?"
subtitle: "A Hypothetical Report for Georgecity University"
authors:
  - name: Kylie King
    email: kk1543@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
bibliography: GIS_Manuscript.bib
title-block-banner: '#27b9f8ff'
title-block-banner-color: "black"
format:
  html:
    df-print: kable
  pdf:
    # https://quarto.org/docs/reference/formats/pdf.html
    link-citations: true
prefer-html: true
---

## Introduction

We're going to start with the story of my SAT experience when we give the presentation.
https://progresslearning.com/news-blog/colleges-dropping-reinstating-act-sat-requirements/
https://www.newamerica.org/education-policy/edcentral/the-sats-not-quite-comeback/
https://www.usnews.com/opinion/articles/2025-10-22/college-testing-act-sat-admissions-meritocracy
https://www.timesnewspapers.com/webster-kirkwoodtimes/sat-act-making-comeback-for-college-admissions/article_77a88f3e-34ab-49c4-b075-58ca91a5ca30.html
https://fairtest.org/overwhelming-majority-of-u-s-colleges-and-universities-remain-act-sat-optional-or-test-blind-score-free-for-fall-2025/

Pretend that you are a lead Data Scientist with an educational
institution.
▪ Your audience are VPs, directors, deans, managers, etc.
▪ It is your job to use the data, evaluate the data and visualize the
data.
• Use Vis method of your choice, more than 1 is ok
▪ It is also your job to use your data analysis results to
communicate the information within the data
▪ Your presentation tells a story!
• It has the beginning, middle, and end
• Use everything that we already learned


Something about how during COVID-19, SAT requirements for college applications were removed or turned optional. I should find a news article about how they are slowly coming back.

We should also have a paragraph about what they're supposed to do, why colleges want standardized testing, and stuff like that.

Then we look at some of the numbers and see if high SAT requirements have anything to do with the percent of students who complete their degrees.


## Exploratory Data Analysis (EDA)

Here we plot the base GIS objects we're analyzing: the location of each **capital city** (in purple) and each **population centroid** (in yellow).

```{r}
#| label: imports
#| echo: false
library(tidyverse) |> suppressPackageStartupMessages()
library(plotly) |> suppressPackageStartupMessages()
library(colorspace) |> suppressPackageStartupMessages()
dark_steelblue <- darken("steelblue4", 0.2)
sat_blue <- '#009bdb'
sat_grey <- "#96999aff"

```

```{r}
#| label: loading_data
#| echo: false
admin_df <- read_csv("data/cleaned_admin_df.csv")
admin_test_df <- read_csv("data/admin_test_full.csv")
df_counts <- admin_test_df |>
  count(Policy_2026) |>
  mutate(pct = n / sum(n) * 100)
```

```{r}
#| label: testing charts
#| echo: false
#testing plotly charts

fig <- plot_ly(admin_df, 
               x = ~SAT_AVG, y = ~C150_4,
               type = 'scatter', mode = 'markers',
               marker = list(color = sat_blue),
               name = 'Schools',
               text = ~paste(
    "School: ", INSTNM,
    "<br>Average SAT: ", SAT_AVG,
    "<br>Completion Rate: ", scales::percent(C150_4, accuracy = 1)
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "Relationship Between SAT Average Scores and 6-Year Completion Rates",
      font = list(size = 18, color = "black", family = "Arial"),
      x = 0.5  # centers the title like hjust = 0.5
    ),
    margin = list(t = 100),
    xaxis = list(
      title = list(text = "SAT Average Score for Accepted Students", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text = "Undergraduate 6-Year Completion Rate", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".0%",
      fixedrange = TRUE
    ),
    legend = list(
      title = list(text = "", font = list(size = 14, color = "black")),
      font = list(size = 12, color = "black")
    ),
    dragmode = "zoom" 
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
#| echo: false
fig
```

```{r}
#| echo: false
fig <- plot_ly(
  df_counts,
  x = ~Policy_2026,
  y = ~pct/100,
  type = 'bar',
  marker = list(
    color = ifelse(df_counts$Policy_2026 == "Test Optional", sat_blue, sat_grey)
  ),
  text = ~paste0(round(pct, 1), "%"),   # Text labels as percentages
  textposition = 'outside',              # Place text above bars
  textfont = list(size = 14, color = "black", family = "Arial")  # Match scatterplot style
) |>
  layout(
    title = list(
      text = "Admissions Test Policies in Nearby Schools for Fall 2026",
      x = 0.5,
      font = list(size = 18, color = "black", family = "Arial")  # Match scatterplot title style
    ),
    xaxis = list(
      title = list(text="Admissions Testing Requirement", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial")
    ),
    yaxis = list(
      title = list(text="Percent of Nearby Schools", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".0%",
      range = c(0, 1)
    ),
    margin = list(t = 90)  # Extra top margin for title
  ) |>
  config(
    staticPlot = TRUE 
  )

```

```{r}
#| echo: false
fig
```

```{r}
#| echo: false

year <- c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
test_takers <- c(1637589, 1715481, 2136539, 2220087, 2198460, 1509133, 1737678, 1913742, 1973891)
test_takers_df <- data.frame(year, test_takers)

fig <- plot_ly(
  test_takers_df,
  x = ~year,
  y = ~test_takers/1e6,
  type = 'scatter',
  mode = 'lines+markers',
  line = list(color = sat_blue, width = 2),
  text = ~paste(
    "Year: ", year,
    "<br>Number of Students:", format(test_takers, big.mark = ",")
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "SAT Test Takers by Year (2016-2024)",
      x = 0.5,
      font = list(size = 16, color = "black", family = "Arial") 
    ),
    xaxis = list(
      title = list(text="Year", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text="Number of Test Takers (millions)", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".1f",
      fixedrange = TRUE
    )
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
#| echo: false
fig
```