---
title: "To Test, or Not to Test?"
subtitle: "DSAN 5900 Homework 1"
authors:
  - name: Kylie King
    email: kk1543@georgetown.edu
    affiliation: Georgetown University
    corresponding: true
df-print: kable
bibliography: GIS_Manuscript.bib
title-block-banner: "#b3d9ff"
title-block-banner-color: "black"
format:
  html:
    df-print: kable
  pdf:
    # https://quarto.org/docs/reference/formats/pdf.html
    link-citations: true
prefer-html: true
---

## Introduction

We're going to start with the story of my SAT experience when we give the presentation.
https://progresslearning.com/news-blog/colleges-dropping-reinstating-act-sat-requirements/
https://www.newamerica.org/education-policy/edcentral/the-sats-not-quite-comeback/
https://www.usnews.com/opinion/articles/2025-10-22/college-testing-act-sat-admissions-meritocracy
https://www.timesnewspapers.com/webster-kirkwoodtimes/sat-act-making-comeback-for-college-admissions/article_77a88f3e-34ab-49c4-b075-58ca91a5ca30.html

Pretend that you are a lead Data Scientist with an educational
institution.
▪ Your audience are VPs, directors, deans, managers, etc.
▪ It is your job to use the data, evaluate the data and visualize the
data.
• Use Vis method of your choice, more than 1 is ok
▪ It is also your job to use your data analysis results to
communicate the information within the data
▪ Your presentation tells a story!
• It has the beginning, middle, and end
• Use everything that we already learned


Something about how during COVID-19, SAT requirements for college applications were removed or turned optional. I should find a news article about how they are slowly coming back.

We should also have a paragraph about what they're supposed to do, why colleges want standardized testing, and stuff like that.

Then we look at some of the numbers and see if high SAT requirements have anything to do with the percent of students who complete their degrees.


## Exploratory Data Analysis (EDA)

Here we plot the base GIS objects we're analyzing: the location of each **capital city** (in purple) and each **population centroid** (in yellow).

```{r}
#| label: imports
library(tidyverse) |> suppressPackageStartupMessages()
library(plotly) |> suppressPackageStartupMessages()
library(colorspace) |> suppressPackageStartupMessages()
dark_steelblue <- darken("steelblue4", 0.2)
sat_blue <- '#009bdb'

theme_eda <- theme(
  # Overall text
  text = element_text(family = "Helvetica", color = "black"),
  
  # Plot background
  plot.background = element_rect(fill = "white", color = NA),
  panel.background = element_rect(fill = "grey98", color = NA),
  
  # Gridlines
  panel.grid.major = element_line(color = "grey85"),
  panel.grid.minor = element_line(color = "grey90"),
  
  # Titles
  plot.title = element_text(size = 20, face = "bold", color = dark_steelblue, hjust = 0.5, margin = margin(b = 10)),
  plot.subtitle = element_text(size = 14, color = "grey30", hjust = 0.5, margin = margin(b = 10)),
  
  # Axis
  axis.title = element_text(size = 14, face = "bold", color = dark_steelblue),
  axis.text = element_text(size = 12, color = "grey20"),
  axis.ticks = element_line(color = "grey70"),
  
  # Legend
  legend.background = element_rect(fill = "grey98", color = NA),
  legend.key = element_rect(fill = "grey98"),
  legend.title = element_text(size = 14, face = "bold", color = dark_steelblue),
  legend.text = element_text(size = 12, color = "grey20"),
  legend.position = "top",
  legend.key.size = unit(1.2, "lines"),
  
  # Facets
  strip.background = element_rect(fill = "steelblue3", color = NA),
  strip.text = element_text(size = 14, face = "bold", color = "white"),
  
  # Margins
  plot.margin = margin(10, 10, 10, 10)
)

# Apply globally
theme_set(theme_eda)

```

```{r}
#| label: loading_data
admin_df <- read_csv("data/cleaned_admin_df.csv")
admin_test_df <- read_csv("data/admin_test_full.csv")
```

```{r}
#| label: testing charts
#testing plotly charts

#fitting a linear model
lm_fit <- lm(C150_4 ~ SAT_AVG, data = admin_df)

# Predict values along the range of SAT_AVG
sat_range <- seq(min(admin_df$SAT_AVG, na.rm = TRUE),
                 max(admin_df$SAT_AVG, na.rm = TRUE),
                 length.out = 100)

pred_df <- data.frame(
  SAT_AVG = sat_range,
  C150_4 = predict(lm_fit, newdata = data.frame(SAT_AVG = sat_range))
)

fig <- plot_ly(admin_df, 
               x = ~SAT_AVG, y = ~C150_4,
               type = 'scatter', mode = 'markers',
               marker = list(color = sat_blue),
               name = 'Schools') %>%
  add_trace(data = pred_df,
            x = ~SAT_AVG, y = ~C150_4,
            type = 'scatter', mode = 'lines',
            line = list(color = 'steelblue3', width = 2),
            name = 'Linear Trend') %>%
  layout(title = list(text = "Relationship Between SAT Average Scores and 6-Year Completion Rates",
                      font = list(size = 16)),
         xaxis = list(title = "SAT Average Score for Accepted Students"),
         yaxis = list(title = "Proportion of Undergrad Students Who Graduate in 6 Years",
                      tickformat = "%"))
```

```{r}
fig
```
