
# Testing all the charts

```{r}
#| label: imports
library(tidyverse) |> suppressPackageStartupMessages()
library(plotly) |> suppressPackageStartupMessages()
library(colorspace) |> suppressPackageStartupMessages()
dark_steelblue <- darken("steelblue4", 0.2)
sat_blue <- '#009bdb'
sat_grey <- "#96999aff"

theme_eda <- theme(
  # Overall text
  text = element_text(family = "Helvetica", color = "black"),
  
  # Plot background
  plot.background = element_rect(fill = "white", color = NA),
  panel.background = element_rect(fill = "grey98", color = NA),
  
  # Gridlines
  panel.grid.major = element_line(color = "grey85"),
  panel.grid.minor = element_line(color = "grey90"),
  
  # Titles
  plot.title = element_text(size = 20, face = "bold", color = dark_steelblue, hjust = 0.5, margin = margin(b = 10)),
  plot.subtitle = element_text(size = 14, color = "grey30", hjust = 0.5, margin = margin(b = 10)),
  
  # Axis
  axis.title = element_text(size = 14, face = "bold", color = dark_steelblue),
  axis.text = element_text(size = 12, color = "grey20"),
  axis.ticks = element_line(color = "grey70"),
  
  # Legend
  legend.background = element_rect(fill = "grey98", color = NA),
  legend.key = element_rect(fill = "grey98"),
  legend.title = element_text(size = 14, face = "bold", color = dark_steelblue),
  legend.text = element_text(size = 12, color = "grey20"),
  legend.position = "top",
  legend.key.size = unit(1.2, "lines"),
  
  # Facets
  strip.background = element_rect(fill = "steelblue3", color = NA),
  strip.text = element_text(size = 14, face = "bold", color = "white"),
  
  # Margins
  plot.margin = margin(10, 10, 10, 10)
)

# Apply globally
theme_set(theme_eda)

```

## Loading data

```{r}
#| label: loading_data
admin_df <- read_csv("cleaned_admin_df.csv")
admin_test_df <- read_csv("admin_test_full.csv")
df_counts <- admin_test_df |>
  count(Policy_2026) |>
  mutate(pct = n / sum(n) * 100)
```


## SAT and Completion scatterplot

```{r}
#| label: testing charts
#testing plotly charts

#fitting a linear model

fig <- plot_ly(admin_df, 
               x = ~SAT_AVG, y = ~C150_4,
               type = 'scatter', mode = 'markers',
               marker = list(color = sat_blue),
               name = 'Schools',
               text = ~paste(
    "School: ", INSTNM,
    "<br>Average SAT: ", SAT_AVG,
    "<br>Completion Rate: ", scales::percent(C150_4, accuracy = 1)
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "Relationship Between SAT Average Scores and 6-Year Completion Rates",
      font = list(size = 18, color = "black", family = "Arial"),
      x = 0.5  # centers the title like hjust = 0.5
    ),
    margin = list(t = 90),
    xaxis = list(
      title = list(text = "SAT Average Score for Accepted Students", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text = "Undergraduate 6-Year Completion Rate", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".0%",
      fixedrange = TRUE
    ),
    legend = list(
      title = list(text = "", font = list(size = 14, color = "black")),
      font = list(size = 12, color = "black")
    ),
    dragmode = "zoom" 
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```


```{r}
fig
```


## Test policy bar chart

```{r}
fig <- plot_ly(
  df_counts,
  x = ~Policy_2026,
  y = ~pct/100,
  type = 'bar',
  marker = list(
    color = ifelse(df_counts$Policy_2026 == "Test Optional", sat_blue, sat_grey)
  ),
  text = ~paste0(round(pct, 1), "%"),   # Text labels as percentages
  textposition = 'outside',              # Place text above bars
  textfont = list(size = 14, color = "black", family = "Arial")  # Match scatterplot style
) |>
  layout(
    title = list(
      text = "Test Policies in Sampled Schools for Fall 2026",
      x = 0.5,
      font = list(size = 18, color = "black", family = "Arial")  # Match scatterplot title style
    ),
    xaxis = list(
      title = list(text="Admission Testing Requirement", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial")
    ),
    yaxis = list(
      title = list(text="Percent of Sampled Schools", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".0%",
      range = c(0, 1)
    ),
    margin = list(t = 90)  # Extra top margin for title
  ) |>
  config(
    staticPlot = TRUE 
  )

fig
```

## Test tester
When SAT scores are standardized, a one–standard deviation increase in a school’s average SAT score is associated with an approximately 15 percentage-point increase in six-year completion rates (p < 0.001). One standard deviation is about 150 SAT score points.

```{r}
value <- 0.15

ggplot() +
  annotate("text", x=0, y=0.7, label = paste0("Colleges have"), size = 6, hjust=0.5) +
  annotate("text", x=0, y=0.5, label = scales::percent(value), size = 20, fontface = "bold", color = sat_blue) +
  annotate("text", x=0, y=0.3, label = paste0("higher 6-year completion rates"), size = 6, hjust=0.5) +
  annotate("text", x=0, y=0.2, label = paste0("per 150 average SAT points above 1170"), size = 5, color="grey50") +
  theme_void() +
  xlim(-0.8, 0.8) + ylim(0, 1)
```

## Test taker line chart

```{r}

year <- c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023, 2024)
test_takers <- c(1637589, 1715481, 2136539, 2220087, 2198460, 1509133, 1737678, 1913742, 1973891)
test_takers_df <- data.frame(year, test_takers)

fig <- plot_ly(
  test_takers_df,
  x = ~year,
  y = ~test_takers/1e6,
  type = 'scatter',
  mode = 'lines+markers',
  line = list(color = sat_blue, width = 2),
  text = ~paste(
    "Year: ", year,
    "<br>Number of Students:", format(test_takers, big.mark = ",")
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "SAT Test Takers by Year (2016-2024)",
      x = 0.5,
      font = list(size = 16, color = "black", family = "Arial") 
    ),
    xaxis = list(
      title = list(text="Year", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text="Number of Test Takers (millions)", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".1f",
      fixedrange = TRUE
    )
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
fig
```


## real text one
```{r}
#| fig-width: 6
#| fig-height: 4
value <- 0.15

ggplot() +
  annotate("text", x=0, y=0.7, label = paste0("The SAT has"), size = 6, hjust=0.5) +
  annotate("text", x=0, y=0.5, label = scales::percent(value), size = 20, fontface = "bold", color = sat_blue) +
  annotate("text", x=0, y=0.3, label = paste0("more predictive power than GPA alone"), size = 6, hjust=0.5) +
  annotate("text", x=0, y=0.2, label = paste0("for mapping first-year college success"), size = 5, color="grey50") +
  theme_void() +
  xlim(-0.8, 0.8) + ylim(0, 1)

```

## cost and sat scores scatterplot

```{r}
#| label: cost chart

fig <- plot_ly(admin_test_df, 
               x = ~SAT_AVG, y = ~COSTT4_A,
               type = 'scatter', mode = 'markers',
               marker = list(color = ~ifelse(Policy_2026 == "Test Required", sat_blue, sat_grey)),
               name = 'Schools',
               text = ~paste(
    "School: ", INSTNM,
    "<br>Average SAT: ", SAT_AVG,
    "<br>Average Cost of Attendance: $", format(COSTT4_A, big.mark = ",")
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "Relationship Between Average SAT Scores and Average Cost of Attendance",
      font = list(size = 18, color = "black", family = "Arial"),
      x = 0.5  # centers the title like hjust = 0.5
    ),
    margin = list(t = 90),
    xaxis = list(
      title = list(text = "SAT Average Score for Accepted Students", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    yaxis = list(
      title = list(text = "Cost of Attendance ($)", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
    legend = list(
      title = list(text = "", font = list(size = 14, color = "black")),
      font = list(size = 12, color = "black")
    ),
    dragmode = "zoom" 
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
fig
```


## Cost and completion scatterplot

```{r}
#| label: ccost and completion

fig <- plot_ly(admin_df, 
               x = ~C150_4, y = ~COSTT4_A,
               type = 'scatter', mode = 'markers',
               marker = list(color = sat_blue),
               name = 'Schools',
               text = ~paste(
    "School: ", INSTNM,
    "<br>Average Cost of Attendance: $", format(COSTT4_A, big.mark = ","),
    "<br>Completion Rate: ", scales::percent(C150_4, accuracy = 1)
  ),
  hoverinfo = 'text'
) |>
  layout(
    title = list(
      text = "Relationship Between School Cost of Attendance and Completion Rate",
      font = list(size = 18, color = "black", family = "Arial"),
      x = 0.5  # centers the title like hjust = 0.5
    ),
    margin = list(t = 90),
    xaxis = list(
      title = list(text = "Undergraduate 6-Year Completion Rate", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      tickformat = ".0%",
      fixedrange = TRUE
    ),
    yaxis =  list(
      title = list(text = "Cost of Attendance ($)", font = list(size = 14, color = "black", family = "Arial")),
      tickfont = list(size = 12, color = "black", family = "Arial"),
      fixedrange = TRUE
    ),
   
    legend = list(
      title = list(text = "", font = list(size = 14, color = "black")),
      font = list(size = 12, color = "black")
    ),
    dragmode = "zoom" 
  ) |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
fig
```


## Sat and admissions rates scatterplot

```{r}
quad_fit <- lm(ADM_RATE ~ SAT_AVG + I(SAT_AVG^2), data = admin_df)

#generating predicitons
x_seq <- seq(min(admin_df$SAT_AVG, na.rm = TRUE),
             max(admin_df$SAT_AVG, na.rm = TRUE),
             length.out = 200)

pred_df <- data.frame(SAT_AVG = x_seq)
pred <- predict(quad_fit, newdata = pred_df, interval = "confidence", level = 0.95)
pred_df$fit <- pred[, "fit"]
pred_df$lwr <- pred[, "lwr"]
pred_df$upr <- pred[, "upr"]

#chart
fig <- plot_ly() |>
  # Scatter points
  add_trace(data = admin_df,
            x = ~SAT_AVG, y = ~ADM_RATE,
            type = 'scatter', mode = 'markers',
            marker = list(color = sat_blue),
            name = 'Schools', text = ~paste(
    "School: ", INSTNM,
    "<br>Average SAT: ", SAT_AVG,
    "<br>Admission Rate: ", scales::percent(ADM_RATE, accuracy = 1)
  ),
  hoverinfo = 'text'
) |>
  # Regression line
  add_trace(data = pred_df,
            x = ~SAT_AVG, y = ~fit,
            type = 'scatter', mode = 'lines',
            line = list(color = sat_grey, width = 2),
            name = 'Line of Best Fit') |>
  # Confidence ribbon
  add_ribbons(data = pred_df,
              x = ~SAT_AVG,
              ymin = ~lwr, ymax = ~upr,
              fillcolor = 'rgba(70,130,180,0.2)',  # semi-transparent steelblue
              line = list(color = 'transparent'),
              name = '95% CI') |>
  layout(title = list(text="Relationship Between SAT Average Scores and Admission Rates",tickfont = list(size = 16, color = "black", family = "Arial")),
         xaxis = list(title = "SAT Average Score for Accepted Students", fixedrange = TRUE, tickfont = list(size = 14, color = "black", family = "Arial")),
         yaxis = list(title = "Admission Rate", fixedrange = TRUE, tickformat = ".0%", tickfont = list(size = 14, color = "black", family = "Arial")),
        dragmode = "zoom") |>
  config(
    modeBarButtonsToRemove = c(
      "select2d",   # box select
      "lasso2d",    # lasso select
      "toggleSpikelines"  # optional extra buttons you might not want
    ),
    displaylogo = FALSE  # removes Plotly logo
  )

```

```{r}
fig
```